---
title: "Visualize CPAN modules with Hilbert curve"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
---

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 8,
    fig.height = 8,
    fig.align = "center"
)
```

There is a nice visualization of CPAN (Perl) modules with the Hilbert curve on http://mapofcpan.org/. In this article, I will
demonstrate how to make the plot with the **HilbertCurve** package.


We first read the list of CPAN modules. The information is in the `02packages.details.txt` file which
can be directly accessed with the following link:

```{r}
df = read.table(url("https://www.cpan.org/modules/02packages.details.txt"), skip = 9)
head(df)
```

The modules are listed in the first column. We also sort it alphabetically.

```{r}
all_modules = sort(df[, 1])
```

We take the "namespace" of the module which is the string before the first "::":

```{r}
ns = gsub("::.*$", "", all_modules)
```

Next we will put every value in `ns` in a Hilbert curve. First let's convert it into an `Rle` object:

```{r}
library(IRanges)
r = Rle(ns)
```

Now in `r`, each element corresponds to an unique namespace in `ns`, and the length or the width of the namespace in `ns` is also calculated. Let's get these values:

```{r}
s = start(r)          # start position of each ns in r
e = end(r)            # end position of each ns in r
w = width(r)          # width of each ns in r
labels = runValue(r)  # corresponding labels
```

Now we can visualize it via the **HilbertCurve** package. We wil highlight the top 30 namespaces with the largest number of modules.

```{r, fig.width = 8, fig.height = 8}
library(HilbertCurve)

ind = order(w, decreasing = TRUE)[1:30]
w_cutoff = w[ ind[30] ]

set.seed(123)  # because we have randomm colors
hc = HilbertCurve(0, length(all_modules), level = 8, mode = "pixel")
hc_layer(hc, x1 = s, x2 = e, 
	col = circlize::rand_color(nrun(r), transparency = ifelse(w >= w_cutoff, 0, 0.8)))
hc_text(hc, x1 = s[ind], x2 = e[ind], labels = labels[ind], 
	gp = gpar(fontsize = 9))
```

```{r}
sessionInfo()
```
